Nice people or potential cooperators when keeping promises

### 2019-08-14 behavioral-promises/behavioral-promises.tex
Update manuscript

\AtBeginDocument{%
  \paperwidth=\dimexpr
    1in + \oddsidemargin
    + \textwidth
    % + \marginparsep + \marginparwidth
    + 1in + \oddsidemargin
  \relax
  \paperheight=\dimexpr
    1in + \topmargin
    + \headheight + \headsep
    + \textheight
    % + \footskip
    + 1in + \topmargin
  \relax
  \usepackage[pass]{geometry}\relax
}

@@

\author{  Said Jiménez  \and  Roberto E. Mercadillo \and  Diego Angeles-Valdez \and  Jairo Muñoz-Delgado \and  Juan J. Sánchez-Sosa \and  Eduardo A. Garza-Villarreal \and  }

@@

\institute{
        Said Jiménez \at
     Department of Psychology, National University of México (UNAM) \\
     \email{\href{mailto:said.ejp@gmail.com}{\nolinkurl{said.ejp@gmail.com}}}  %  \\
 %           \emph{Present address:} of F. Author  %  if needed
    }

@@

\begin{abstract}
Promises increase cooperation between non-genetically related
individuals, however, lies and betrayals occur daily in close social
relationships. Despite this, the effect of social closeness on the
decision to keep or break a promise has not been studied. We conducted
an experiment in which subjects could freely decide if they broke or
kept their promise to three partners with different levels of social
closeness: null (computer), low (strange) and high (friend). If subjects
keep their word with the computer, it provides evidence in favor of
intrinsic motivation (comply to do the correct thing), otherwise
extrinsic motivation would be supported (comply to facilitate future
cooperation). Using Hierarchical Bayesian Modeling, we find that if
social closeness increases, the probability of breaking promises
decreases monotonically, the evidence ratio in favor to the hypothesis
stranger \textless{} computer was 7.47, and friend \textless{} stranger
was 8.64. Furthermore, using Bayesian Cumulative Modeling we find that
subjects were very consistent between their promises and the subsequent
decisions they make. Although is more probable that our subjects keep
their promises to any of the three partners, there is a proportion of
transgressions to the computer, whose inference excludes zero with a
95\% posterior probability. Also, subjects never broke the promise to
their friends. Our results indicate that subjects are predominantly
``nice people'' because they keep their promises even to partners
without social closeness, however, if the subjects are not certain that
their partner will also be a ``potential cooperator'', dishonesty
emerges in the form of broken promises.
\\
\keywords{
        Deception \and
        Hierarchical Bayesian Models \and
        Bayesian Cumulative Models \and
        Social closeness \and
        Promises \and
    }

@@

Human are social animals and we get survival advantages from being such.
Historically, belonging to a social group has provided protection and
guaranteed access to resources such as food, and currently, there has
been evidence that belonging to a social group has a positive effect on
longevity, physical and mental health (Tough, Siegrist, and Fekete 2017;
Holt-Lunstad 2018). The relationship between social closeness and
cooperation is key to the formation of large groups of individuals
without a genetic relationship, and such groups form the basis of
communities, societies, and nations, arguably constituting one of the
most fundamental conditions for human survival (Fehr and Fischbacher
2003; Fehr and Schurtenberger 2018). An important requirement for
cooperation between individuals is communication. A meta-analysis of 45
studies reported a large positive effect (\(d = 1.01\)) of communication
in cooperation regardless of the communication medium (Balliet 2010). In
parallel, it is known that after communication sessions between
completely unknown individuals there is an increase in subjective social
closeness (Aron et al. 1997). Thus, these findings suggest a positive
relationship between communication, cooperation, and social closeness.

One of the forms of communication that have received the most attention
in psychological game theory is ``the promise''. In an interaction
between sender and recipient, it is believed that promises influence
beliefs of the recipient, generating trust and cooperation (Charness and
Dufwenberg 2006; Vanberg 2008). However, there are situations where
trust is betrayed, promises are broken or people deceived. The General
Social Survey (GSS) in 2018 showed that approximately 11\% of people
surveyed responded that they have had sex with someone different to
their partner while they are married (Smith et al. 2018). In economic
and social terms, the United Nations Organization has stressed that
deception (corruption) has a global cost of at least 5 percent of the
annual gross domestic product (GDP), and it is a generator of conflict
and instability in all nations (UN 2018).

Despite the widespread presence of interactions between human belonging
to the same group, it is not known how social closeness between
participants can affect keeping or breaking promises. This is important
because it has been pointed out that lies and betrayals of trust occur
daily in close relationships, such as friends, workmates, or classmates.
And, their consequences can be serious such as loss of the relationship,
work or damage to reputation (DePaulo et al. 2004). In the laboratory,
these transgressions to trust have been explicitly studied in
experiments where subjects have incentives to lie or not keep their
promises. However, the vast majority of these studies have been
conducted in people who do NOT know each other (Gneezy 2005; Gneezy,
Rockenbach, and Serra-Garcia 2013; Baumgartner et al. 2009; Baumgartner,
Gianotti, and Knoch 2013; Charness and Dufwenberg 2006; Mazar, Amir, and
Ariely 2008; Fischbacher and Föllmi-Heusi 2013). Some studies address
deception in close interpersonal relationships using self-report
measurements, however, they have limitations due to social desirability
bias (DePaulo and Kashy 1998; DePaulo et al. 2004) or rather explore the
development of deception detection skills in same-sex friends (Anderson,
DePaulo, and Ansfield 2002).

In this study, we wanted to understand the effects of three partners
with different levels of social closeness on keeping or breaking
promises, as well as on cooperation, using an experimental trust game.
We analyzed it in light of the two main motivations that have been
pointed out in the literature regarding keeping promises which are:
instrumental and intrinsic (Baumgartner et al. 2009):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The \emph{instrumental} suggests that promises are kept to make future
  cooperation easier.
\item
  The \emph{intrinsic} mentions that the promises are kept to do what is
  morally right.
\end{enumerate}

In our study, subjects perform a standard trust game in pairs with three
phases: \emph{first}, the trustee makes a promise to pay half of his
earnings regardless of who his investor is; \emph{second}, the investor
receives the promise and decides if he invests his initial budget;
\emph{third}, in case the investor has given the budget, the trustee
faces the decision to pay or not to pay half of his earnings. To
evaluate the effect of social closeness, our subjects participate in the
role of trustee in front of three partners with different levels of
social closeness (null, low and high) in the role of the investor: a
computer, a stranger, and a friend. The manipulations mentioned allow us
to evaluate several hypotheses, the first one was proposed a priori and
is derived from a larger research project (you can check the
preregistration of the hypothesis here: \url{https://osf.io/u97fd}),
while the following are exploratory:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Social closeness will reduce the decision of breaking the promise.
  According to the \emph{instrumental} motivation, we expect that
  subjects keep their promises with friends for the purpose of
  facilitating future cooperation, which can be extended even beyond the
  trials in the experiment. We also expect, although to a lesser extent,
  that the subjects keep promises to strangers, with the purpose of
  facilitating cooperation at least during the trials during the
  experiment. Finally, we anticipate that the participants break the
  promises to the computer because it is a partner without social
  closeness and they could not ensure cooperation in future trials. It
  should be noted that if the participants keep the promises to the
  computer, evidence would be given in favor of \emph{intrinsic}
  motivation.
\item
  There will be an effect of social closeness on cooperation, regardless
  of promises. We expect that there is more probability of paying the
  friend than the stranger and more probability of paying the stranger
  than the computer. According to the \emph{instrumental motivation},
  the cooperation will be greater for partners with whom the subject
  anticipates greater cooperation in the future (friend \textgreater{}
  stranger \textgreater{} computer).
\item
  Finally, we will obtain two subsamples according to the cooperation
  rate of the subjects as was done in another study (Baumgartner et al.
  2009). However, what in the aforementioned study was classified as a
  group of ``honest'' and ``dishonest'', we will show that in our sample
  it is not supported because if the groups differ in their payment rate
  it will be due to the difference in their level of commitment
  expressed by promises.
\end{enumerate}

@@

We recruited 45 subjects (15 male) from the National Autonomous
University of Mexico, the age range from 19 to 33 years and their
minimum educational level were bachelor. Subjects were asked to come to
the study with a ``close'' friend, who fulfilled the following
characteristics: 1) he/she was matched by sex, 2) did not have a family
bond and 3) was not a person with a sentimental or sexual relationship.
From this sample, 30 subjects (15 men) performed the task in a magnetic
resonance imaging (MRI) scanner, however, their image data was not
analyzed for the present work.

@@

All subjects performed an adaptation of the trust game with promises,
using hypothetical monetary rewards (Baumgartner et al. 2009). The
variation with respect to the original task is that in our experiment
three partners were presented with three levels of social closeness in
the role of the investor: computer (no closeness), strange (low
closeness) and friend (high closeness), while the participant acted as
the trustee. The task was programmed in PsychoPy2 version 1.84.2 (Peirce
et al. 2019; Peirce 2008) and consisted of 24 trials between two
players: trustee and investor. The trustee originally has 0 Mexican
pesos and the investor has 2 pesos, the investor is presented with the
opportunity to give his money to the trustee or keep it. If investor
gives his money, it is multiplied by 5 pesos, so that the trustee has 10
pesos. Finally, the trustee decides to pay half to the investor or keep
the 10 pesos. The structure described is repeated in 24 trials, however,
in 4 of the trials the trustee can send a promise to the investor, the
promises were that always, mostly, sometimes, or never would pay back.
Each promise was valid for three trials so that 12 of the trials have
the effect of the promise and the other 12 do not. Each partner performs
the role of investor in 8 of 24 total trials, however, both the promises
and social closeness conditions were presented to the trustee in
pseudorandom order.

Since our main interest was the trustee's behavior, investors' decisions
were programmed \emph{a priori} to give their amount in 6 trials and in
2 did not randomly. The covert story for all our subjects was that they
would be playing in real-time with their friend (whom they brought to
the study), the stranger (was told he would be another same-sex unknown
person) and the computer. After the study ended, the participants were
debriefed about our own deception. A diagram of the chronology of the
experimental task with the duration of each phase in seconds is shown in
Figure 1.

@@

Fixation phase consisted only of a period in which the subject paid
attention without performing any particular behavior. Then,
\textbf{promises phase} in part A1 indicated that in the following three
trials they could decide without the effect of the promise. In A2, the
subject was asked to decide between 1) always, 2) mostly, 3) sometimes
or 4) never pay back. This pattern was repeated. In the
\textbf{anticipation/assignation phase}, subjects were told who their
partner was for that trial (computer, stranger or friend) and was given
the message that their partner was making their decision. Later in the
\textbf{decision phase}, the subject was informed if his partner
invested his \$2 or not, he was reminded of his promise level (if they
were trails with promises) and, in case the partner had invested, he was
asked to decide whether to pay back or not. Finally, payments for each
trial were shown in the \textbf{feedback phase} and the sequence was
repeated until the task finished.

@@

\caption{Pay back proportion by partners and promises levels}\label{fig:fig2}
\end{figure}

@@

\caption{Posterior probability of Promise Breaking}\label{fig:fig3}
\end{figure}

@@

\caption{Posterior estimates in logit scale}\label{fig:fig4}
\end{figure}

@@

\caption{Posterior predictive over pay back rate}\label{fig:fig5}
\end{figure}

@@

\caption{Posterior predictive over pay back rate, Social Closeness Varying effects by Promise levels}\label{fig:fig6}
\end{figure}

@@

\caption{Promise choices by group}\label{fig:fig7}
\end{figure}

@@

To our knowledge, this is the first experimental study that incorporates
social closeness as a predictor of the decision to break promises with
socially relevant partners. In previous studies, the participants remain
anonymous during the course of the tasks (Baumgartner et al. 2009),
their measurements to evaluate transgressions are self-reported (DePaulo
and Kashy 1998; DePaulo et al. 2004), they do not directly quantify the
breaking of promises (Vanberg 2008; Charness and Dufwenberg 2006), or as
we will discuss later, they use heterogeneous measures of social
closeness (Glaeser et al. 2000). Our experiment is the first one to
include a partner without social closeness (computer). Our data suggest
that the mere fact of considering that subjects play with a human
diminishes the probability of breaking the promise. It is more likely to
break the promise to the computer than to the stranger, even though that
partner was not known. Also, null social closeness is important because
it allows us to explore the intrinsic motivation to keep the promises.
If humans keep their word because it is morally correct, we would
anticipate that the probability of breaking a promise was very low in
the three partners and, particularly, the inference regarding the
probability of breaking the promise to the computer would include zero.
However, we found that, although low, the probability was greater than
zero, which provides evidence that seems to contradict, at least in
degree, the hypothesis of intrinsic motivation. We cannot exclude the
possibility that the subjects keep promises mainly due to moral
motivation because, even in the case of the computer, the subjects kept
a large proportion. We can, however, exclude that this is their only
motivation.

On the other hand, high social closeness allows us to explore the
instrumental motivation to keep promises (Baumgartner et al. 2009). In
our study, it was more likely to break the promise to the stranger than
to the friend. If humans keep their word to facilitate cooperation in
the future, we would anticipate that with a partner of high social
closeness (with whom they are very likely to cooperate, even after the
experiment) there would be less of a chance to break the promise. In
fact, the promise to friends was not broken on any occasion, hence the
inference in this case, was met.

The findings concerning motivations could be explained by the social
norm called ``conditional cooperation'', which indicates that the belief
that other people cooperate at high levels also induces high levels of
cooperation (Fehr and Schurtenberger 2018). Thus, the lack of certainty
regarding the decisions that the computer would make could explain the
probability of breaking the promise that exists towards this partner.
Similarly, the information that the subjects have regarding as their
friend's behavior -even before the experiment- could explain the high
levels of cooperation and keeping of promises towards this partner.

Cooperation has been studied with diverse experimental manipulations. In
one study, the contribution in a public goods game was evaluated during
several trials in a group made up of the same individuals (partners),
and compared to another group of new different subjects for each trial
(strangers) (Croson 1996). In the condition of partners the cooperation
was greater because it is a stable group compared to the group of
strangers. The mentioned study does not give details regarding the
recruitment of participants, so we could assume that even in the
condition of partners these are individuals who do not consider
themselves socially close.

Another study showed an increase in cooperation in a trust game when
individuals were socially close (Glaeser et al. 2000). In this study,
subjects knew each other and the researchers carefully measured several
variables regarding their social connection.However, some individuals
who arrived together were allowed to perform the task between themselves
and others were paired with unclear criteria. The above allows social
closeness between partners to be heterogeneous, allows for romantic
relationships and assumes that two individuals who come together to
class are considered close to each other. Also, since it is a study of
one-shot, it excludes the possibility of evaluating how the same subject
varies his behavior based on different levels of social closeness.

To try to homogenize social closeness between our subjects and their
partners, they performed the task with a same-sex friend considered to
be close, emphasizing their companion could not be a romantic or sexual
partner. Bayesian analysis of the IOS scale showed evidence in favor
that the degree of social closeness between 30 of our 45 subjects and
their friends was the same. Results of the two mentioned studies also
found that social closeness enhance cooperation (Croson 1996; Glaeser et
al. 2000), in the future would be valuable to homogenize both the
performed tasks and the statistical procedure used, to establish
similarity in effects magnitudes.

Our methodological contribution to the field is the use of Bayesian
inference tools: Hierarchical Models and Cumulative Models. The
Hierarchical Models, allowed us to model the effect of the experimental
conditions on the individual response. These models assume that lower
level observations (e.g.~decisions of each individual) are nested in
higher level units (e.g.~individual subjects). Within-subjects designs
have traditionally been analyzed with repeated measures AN(C)OVA,
however, Hierarchical Models grant the advantages of naturally dealing
with unbalanced data, include categorical and/or numerical predictors,
explicitly incorporate individual variability, among others (Vuorre and
Bolger 2018; Gelman and Hill 2006; McElreath 2018; Bürkner 2017).

In our case, the Hierarchical Model allowed us to capture how the
experimental conditions affected the decisions of each subject and the
differences between them. For example, there are obviously motivated
instrumental subjects such as number 8 or 12 who are very sensitive to
the identity of their investors and modify their cooperation under
social closeness. At the same time, there were notably intrinsic
subjects such as 9 and 10 who cooperate all the time regardless of who
their partners are. The Hierarchical Model naturally includes this
information for the estimation of population effects, which, if not
considered, would lead to inaccurate inferences(Bürkner 2017; Gelman and
Hill 2006; McElreath 2018).

On the other hand, the Cumulative Model allowed us to capture the
strength of the commitment expressed through promises. A very frequent
problem that has been pointed out recently is that analyzing ordinal
data with methods that assume that observations are metric can lead to
serious inference errors (Liddell and Kruschke 2018). The Cumulative
Models assumes that the observed responses come from the categorization
of a continuous latent variable (Bürkner and Vuorre 2019). In our study,
we could identify that groups with different payment proportions, which
in other studies have been called ``dishonest'' and
``honest''(Baumgartner et al. 2009; Baumgartner, Gianotti, and Knoch
2013), differ in the commitment they express through their promises.
Considering that the tools to perform the Cumulative Models are
relatively recent, in future studies, it would be valuable to also use
the new alternatives for inference.

Although the Hierarchical and Cumulative models are not tools
exclusively for Bayesian inference, their application from this approach
represents several advantages compared to frequentist statistics.
Classic problems such as multiple comparisons, the decision of when to
stop collecting subjects (stopping rule), or use of planned comparisons
versus post hoc, are not factors that affect the Bayesian approach
(Dienes 2011). In Bayesian inference, hypothesis testing makes formal
use of probability to express the plausibility of theories, and in our
case, we were able to obtain evidence ratios regarding the extent of our
data support hypotheses, regardless if these were proposed \emph{a
priori} or \emph{post hoc}.

>>>>

@@

\author{  Said Jiménez \and  Kamila E. Sip \and  Roberto E. Mercadillo \and  Diego Angeles-Valdez \and  Jairo Muñoz-Delgado \and  Juan J. Sánchez-Sosa \and  Eduardo A. Garza-Villarreal \and  }

@@

\institute{
        Said Jiménez \at
     Department of Psychology, National University of México (UNAM) \\
     \email{\href{mailto:said.ejp@gmail.com}{\nolinkurl{said.ejp@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Eduardo A. Garza-Villarreal \at
     Institute of Neurobiology, UNAM \\
     \email{\href{mailto:egarza@gmail.com}{\nolinkurl{egarza@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed

@@

\begin{abstract}
Promises increase cooperation between non-genetically related
individuals, whereas lies and betrayals occur daily even in close social
relationships. Despite this, the effect of social closeness on the
decision to keep or break a promise has not been thoroughly studied. We
conducted an experiment in which subjects could freely decide if they
broke or kept a promise to three partners with different levels of
social closeness: null (computer), low (strange) and high (friend). If
subjects kept their word with the computer, it provides evidence in
favor of intrinsic motivation (comply to do the correct thing), while
extrinsic motivation would be supported (comply to facilitate future
cooperation). Using Hierarchical Bayesian Modeling, we found that as
social closeness increases, the probability of breaking promises
decreases monotonically. The evidence ratio in favor of the hypothesis
stranger \textless{} computer was 7.47, and friend \textless{} stranger
was 8.64. Furthermore, using Bayesian Cumulative Modeling we found that
subjects were very consistent between their promises and their choices.
Although it was more probable that our subjects kept their promises to
any of the three partners, there was a proportion of transgressions to
the computer, whose inference excludes zero with a 95\% posterior
probability. Also, our subjects never broke the promise to their
friends. Our results suggest that subjects were predominantly ``nice
people'' because they kept their promises even to partners without
social closeness. However, if the subjects were not certain that their
partner will also be a ``potential cooperator'', dishonesty emerged in
the form of broken promises.
\\
\keywords{
        Deception \and
        Hierarchical Bayesian Models \and
        Bayesian Cumulative Models \and
        Social closeness \and
        Promises \and
    }

@@

Humans are social animals and we obtain survival advantages from proper
socialization. Historically, belonging to a social group has provided
protection and guaranteed access to resources such as food, and
currently, there is evidence that belonging to a social group has a
positive effect on longevity, physical and mental health (Tough,
Siegrist, and Fekete 2017; Holt-Lunstad 2018). The interaction between
social closeness and cooperation seems to be a key feature to the
formation of large groups of individuals without a genetic relationship,
and such groups form the basis of communities, societies, and nations,
arguably constituting one of the most fundamental conditions for human
survival (Fehr and Fischbacher 2003; Fehr and Schurtenberger 2018). A
recent meta-analysis showed that communication, regardless of the
medium, is the most important element for cooperation between
individuals (Balliet 2010). In parallel, there is evidence that
following communication sessions between completely unknown individuals,
there is an increase in subjective social closeness(Aron et al. 1997).
These findings suggest the importance of the relationship between
communication, cooperation, and social closeness.

One of the forms of communication that have received the most attention
in psychological game theory is ``the promise''. In an interaction
between a sender and a recipient, it is believed that the sender's
promises influence the recipient's beliefs, influencing their ability to
trust and cooperate (Charness and Dufwenberg 2006; Vanberg 2008).
Commonly, however, there are situations where promises are broken, trust
is betrayed and people are deceived. The General Social Survey (GSS) in
2018 showed that approximately 11\% of people surveyed responded that
they have had sexual intercourse with someone different to their partner
while they were married (Smith et al. 2018). In economic and social
terms, the United Nations Organization has stressed that deception
(corruption) has a global cost of at least 5 percent of the annual gross
domestic product (GDP), and it is a generator of conflict and
instability in all nations (UN 2018). Hence, breaking promises may
greatly affect our own social groups and society itself.

Despite the widespread presence of interactions between human belonging
to the same group, it is not known how social closeness between
participants can affect keeping or breaking promises. This is important
because it has been pointed out that lies and betrayals of trust occur
daily even in close relationships such as friends, coworkers, or
classmates, and their consequences can be serious, from loss of the
relationship to work loss and damage to reputation (DePaulo et al.
2004). The two main motivations that have been pointed out in the
literature regarding keeping promises are: instrumental and intrinsic
(Baumgartner et al. 2009). The instrumental suggests that promises are
kept to make future cooperation easier while the intrinsic suggests that
the promises are kept to do what is ``morally right''. In the
laboratory, the betrayals of trust have been explicitly studied in
experiments where subjects have incentives to lie and break promises.
However, the vast majority of studies have been conducted in people
without any social closeness (strangers) (Gneezy 2005; Gneezy,
Rockenbach, and Serra-Garcia 2013; Baumgartner et al. 2009; Baumgartner,
Gianotti, and Knoch 2013; Charness and Dufwenberg 2006; Mazar, Amir, and
Ariely 2008; Fischbacher and Föllmi-Heusi 2013). Some studies have
addressed deception in close interpersonal relationships using
self-report measurements or rather explore the development of deception
detection skills in same-sex friends, with clear limitations due to the
social desirability bias (DePaulo and Kashy 1998; DePaulo et al. 2004;
Anderson, DePaulo, and Ansfield 2002).

In this study, we wanted to understand the effects of social closeness
on keeping or breaking promises and cooperation, by means of an
experimental trust game with three partners with different levels of
social closeness. For this, subjects performed a standard trust game in
pairs with three phases: 1) the trustee makes a promise to pay half of
his earnings regardless of who his/her investor; 2) the investor
receives the promise and decides if he/she invests his/her initial
budget; 3) in case the investor has given the budget, the trustee faces
the decision to pay or not to pay half of his earnings (keep or break
promise). To evaluate the effect of social closeness, our subjects
participate in the role of trustee with three partners with different
levels of social closeness (zero, low and high) in the role of the
investor: a computer, a stranger, and a close friend. Our main
pre-registered hypothesis was:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Social closeness will reduce the decision of breaking the promise.
  According to the instrumental motivation, we expect that subjects keep
  their promises with friends for the purpose of facilitating future
  cooperation, which probably can be extended even beyond the trials in
  the experiment. We also expect, although to a lesser extent, that the
  subjects keep promises to strangers, with the purpose of facilitating
  cooperation during the experiment. Finally, we anticipate that the
  participants break the promises to the computer because it is a
  partner without any social closeness and they could not ensure
  cooperation in future trials. It should be noted that if the
  participants keep the promises to the computer, evidence would be
  given in favor of intrinsic motivation.
\end{enumerate}

Our secondary exploratory hypotheses were:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\item
  There will be an effect of social closeness on cooperation, regardless
  of promises. According to the instrumental motivation, the payment
  will be greater for partners with whom the subject anticipates greater
  cooperation in the future (friend \textgreater{} stranger
  \textgreater{} computer).
\item
  We hypothesize that the subjects will cooperate according to their
  degree of commitment expressed through promises.
\end{enumerate}

@@

We recruited 45 subjects (15 male) from the National Autonomous
University of Mexico, age range from 19 to 33 years, their minimum
educational level were bachelor and right handed (for MRI). Subjects
were asked to come to the study with a ``close'' friend, who fulfilled
the following criteria: 1) he/she was matched by sex, 2) did not have a
family bond and 3) was not a person with a sentimental or sexual
relationship. From this sample (n = 45), 30 subjects (15 men) performed
the task in a magnetic resonance imaging (MRI) scanner, however, their
image data was not analyzed for the present work.

@@

All subjects performed an adaptation of the trust game with promises,
using hypothetical monetary rewards (Baumgartner et al. 2009). The
variation with respect to the original task was that, in our experiment,
three partners with three levels of social closeness were presented in
the role of the investor: computer (no closeness), strange (low
closeness) and friend (high closeness), while the participant acted as
the trustee. The task was programmed in PsychoPy2 version 1.84.2 (Peirce
et al. 2019; Peirce 2008) and consisted of 24 trials between two
players: trustee and investor. The trustee originally had 0 Mexican
pesos and the investor had 2 pesos, the investor was presented with the
opportunity to give his/her money to the trustee or keep it. If the
investor gave his/her money, it was multiplied by 5 pesos, so that the
trustee had 10 pesos. Finally, the trustee had to decide if he/she
wanted to pay half to the investor or keep the 10 pesos. The structure
described was repeated in 24 trials, however, in 4 of the trials the
trustee could send a promise to the investor, the promises were that
he/she would: always, mostly, sometimes, or never pay back. Each promise
was valid for three trials, so that 12 of the trials had the effect of
the promise and the other 12 did not. Each partner performed the role of
investor in 8 of 24 total trials, however, both the promises and social
closeness conditions were presented to the trustee in pseudorandom
order.

Since our main interest was the trustee's behavior and to avoid
unbalanced responses from investors, the investor was always the
computer although the trustee was shown to have 3 types of investors,
and the investors' decisions were programmed \emph{a priori} to randomly
give their amount in 6 trials and withhold the amount in 2 trials. The
covert story for all our subjects was that they would be playing in
real-time with their friend (whom they brought to the study), the
stranger (was told it would be another same-sex unknown person) and the
computer. After the study ended, the participants were debriefed about
our deception. A diagram of the the experimental task is shown in Figure
1.

@@

Promise phase in part A1 indicated that in the following three trials
they could decide without the effect of the promise. In A2, the subject
was asked to decide between 1) always, 2) mostly, 3) sometimes or 4)
never pay back. This pattern was repeated. In the
anticipation/assignation phase, subjects were told who their partner was
for that trial (computer, stranger or friend) and was given the message
that their partner was making their decision. Later in the decision
phase, the subject was informed if his partner invested his \$2 or not,
he was reminded of his promise level (if they were trails with promises)
and, in case the partner had invested, he was asked to decide whether to
pay back or not. Finally, payments for each trial were shown in the
feedback phase and the sequence was repeated until the task finished.

@@

\caption{Pay back proportion by partners and promises levels, the x-axis represents all decisions made to each partner, the different shades of blue shows the proportions in which the subjects paid back or not. The y-axis presents how these proportions vary according to the promises made.}\label{fig:fig2}
\end{figure}

@@

\caption{Inference regarding the probability of breaking the promise depending on social closeness. The x-axis shows the posterior probability of breaking the promise, and the y-axis indicates the three partners. The dotted line shows the probability that breaking the promise would be zero.}\label{fig:fig3}
\end{figure}

@@

\caption{Posterior estimates in logit scale. We show posteriors distributions of the hierarchical Bayesian model coefficients for the effect of experimental conditions on the decision to cooperate. The x-axis shows the magnitude of the logit scale coefficients, and the y-axis shows the experimental conditions associated with that effect.}\label{fig:fig4}
\end{figure}

@@

\caption{Posterior predictive over pay back rate. Correspondence between payment proportions and model predictions depending on the game partner for the first twelve subjects in the sample. The x-axis pointed the three playmates when the subjects had made a promise, while the y-axis indicates the proportion of decisions in which subjects payback.}\label{fig:fig5}
\end{figure}

@@

\caption{Posterior predictive over pay back rate, social closeness varying effects by promise levels. Correspondence between payment proportions and model predictions depending on the game partner at the group level. The x-axis pointed the three playmates when the subjects had made a promise, while the y-axis indicates the proportion of decisions in which subjects payback.}\label{fig:fig6}
\end{figure}

@@

\caption{Promise choices by group. We show posterior probability estimates regarding the level of promise depending on the group. The Low group (which paid 58/100 of the trials) is more likely to promise mostly, while the High group (which paid 83/100 of the trials) is more likely to promise always.}\label{fig:fig7}
\end{figure}

@@

To our knowledge, this is the first experimental study that incorporates
social closeness as a predictor of the decision to break promises with
socially relevant partners. In previous studies, the participants remain
anonymous during the course of the tasks (Baumgartner et al. 2009),
their measurements to evaluate transgressions are self-reported (DePaulo
and Kashy 1998; DePaulo et al. 2004), they do not directly quantify the
breaking of promises (Vanberg 2008; Charness and Dufwenberg 2006), or as
we will discuss later, they use heterogeneous measures of social
closeness (Glaeser et al. 2000). Our results suggest that the mere fact
of considering that subjects play with a human diminishes the
probability of breaking the promise. It is more likely to break the
promise to the computer than to the stranger, even though that partner
was not known. Also, zero social closeness is important to include
because it allows us to explore the intrinsic motivation to keep the
promises. If humans keep their word because it is morally correct, we
would anticipate that the probability of breaking a promise was very low
in the three partners and, particularly, the inference regarding the
probability of breaking the promise to the computer would include zero.
However, we found that, although low, the probability was greater than
zero, which provides evidence that seems to contradict, at least in some
degree, the hypothesis of intrinsic motivation. This means that we
cannot exclude the possibility that the subjects keep promises mainly
due to moral motivation because, even in the case of the computer, the
subjects kept a large proportion of promises. We can, however, exclude
that this is their \emph{only} motivation.

High social closeness allows us to explore the instrumental motivation
to keep promises (Baumgartner et al. 2009). In our study, subjects were
more likely to break the promise to the stranger than to the friend. If
humans keep their word to facilitate cooperation in the future, we would
anticipate that there would be less of a chance to break the promise
with a partner of high social closeness (with whom they are very likely
to cooperate, even after the experiment) and, in fact, the promise to
friends was not broken on any occasion. The findings concerning
motivations could be explained by the social norm called ``conditional
cooperation'', which indicates that the belief that other people
cooperate at high levels also induces high levels of cooperation (Fehr
and Schurtenberger 2018). Thus, the uncertainty regarding the computer's
decisions could explain the probability of breaking the promise.
Similarly, the information that the subjects have regarding their
friend's behavior, even before the experiment, could explain the high
levels of cooperation and keeping of promises towards the friend.

Cooperation has been studied with diverse experimental manipulations. In
one study, the contribution in a ``public goods game'' was evaluated
during several trials in a group made up of the same individuals
(partners), and compared to another group of new different subjects for
each trial (strangers) (Croson 1996). In the condition of partners, the
cooperation was greater because it was a stable group compared to the
group of strangers. The mentioned study does not give details regarding
the recruitment of participants, thus we could assume that even in the
condition of partners these were individuals who did not consider
themselves socially close. Another study showed an increase in
cooperation in a trust game when individuals were socially close
(Glaeser et al. 2000). In this study, subjects knew each other and the
researchers carefully measured several variables regarding their social
connection such as the number of friends in common and years of
relationship. They found greater cooperation depending on the degree of
social connection. However, the study did not control for social
closeness as some individuals who arrived together were allowed to
perform the task together whereas others were paired differently. This
limitation allowed social closeness between partners to be
heterogeneous, allowed for romantic relationships and assumed that two
individuals who arrived together to class were considered close to each
other. To try to homogenize social closeness between our subjects and
their partners, they performed the task with a same-sex friend
considered to be close, emphasizing their companion could not be a
romantic or sexual partner. Bayesian analysis of the IOS scale showed
evidence in favor that the degree of social closeness between 30 of our
45 subjects and their friends was the same. Although results of the two
mentioned studies also found that social closeness enhance cooperation
(Croson 1996; Glaeser et al. 2000), it would be valuable to homogenize
both the performed tasks and the statistical procedure used to establish
similarity in effects magnitudes.

Our methodological contribution to the field is the use of Bayesian
inference tools: Hierarchical Models and Cumulative Models. The
Hierarchical Models, allowed us to model the effect of the experimental
conditions on the individual response. These models assume that lower
level observations (e.g.~decisions of each individual) are nested in
higher level units (e.g.~individual subjects). Within-subjects designs
have traditionally been analyzed with repeated measures AN(C)OVA,
however, Hierarchical Models grant the advantages of naturally dealing
with unbalanced data, include categorical and/or numerical predictors,
explicitly incorporate individual variability, among others (Vuorre and
Bolger 2018; Gelman and Hill 2006; McElreath 2018; Bürkner 2017). In our
case, the Hierarchical Model allowed us to capture how the experimental
conditions affected the decisions of each subject and the differences
between them. For example, there were obviously motivated instrumental
subjects such as number 8 or 12 who are very sensitive to the identity
of their investors and modify their cooperation under social closeness.
At the same time, there were notably intrinsic subjects such as 9 and 10
who cooperate all the time regardless of who their partners are. The
Hierarchical Model naturally includes this information for the
estimation of population effects, which, if not considered, would lead
to inaccurate inferences (Bürkner 2017; Gelman and Hill 2006; McElreath
2018). The Cumulative Model allowed us to capture the strength of the
commitment expressed through promises. A very frequent problem that has
been pointed out recently is that analyzing ordinal data with methods
that assume that observations are metric can lead to serious inference
errors (Liddell and Kruschke 2018). The Cumulative Models assumes that
the observed responses come from the categorization of a continuous
latent variable (Bürkner and Vuorre 2019). In our study, we could
identify that groups with different payment proportions, which in other
studies have been called ``dishonest'' and ``honest'' (Baumgartner et
al. 2009; Baumgartner, Gianotti, and Knoch 2013), differ in the
commitment they express through their promises. Considering that the
tools to perform the Cumulative Models are relatively recent, in future
studies, it would be valuable to also use the new alternatives for
inference. Although the Hierarchical and Cumulative models are not tools
exclusively for Bayesian inference, their application from this approach
represents several advantages compared to frequentist statistics.
Classic problems such as multiple comparisons, the decision of when to
stop collecting subjects (stopping rule), or use of planned comparisons
versus post hoc, are not factors that affect the Bayesian approach
(Dienes 2011). In Bayesian inference, hypothesis testing makes formal
use of probability to express the plausibility of theories, and in our
case, we were able to obtain evidence ratios regarding the extent of our
data support hypotheses, regardless if these were proposed \emph{a
priori} or \emph{post hoc}.
